Algorithms for Longest Common Extensions

### 2011-08-20 presentation/presentation.tex
m

\section{Introduction}
\subsection{The LCE Problem}
\begin{frame}{The LCE Problem and the \proc{DirectComp} algorithm}
    \begin{block}{Input}
        \begin{itemize}
            \item $s=abbababba$
            \item $(i,j)=(4,6)$
        \end{itemize}
    \end{block}
    \begin{block}{The \proc{DirectComp} algorithm}
        %\setlength{\unitlength}{1cm}
        \definecolor{matchcolor}{rgb}{.6,1,.6}
        \definecolor{mismatchcolor}{rgb}{1,.6,.6}
        \begin{picture}(300,50)
            \put(0, 30){\makebox(20, 20){$s=$}}
            \put(20, 30){\framebox(20, 20){a}}
            \put(40, 30){\framebox(20, 20){b}}
            \put(60, 30){\framebox(20, 20){b}}
            \put(80, 30){\framebox(20, 20){a}}
            \put(100, 30){\framebox(20, 20){b}}
            \put(120, 30){\framebox(20, 20){a}}
            \put(140, 30){\framebox(20, 20){b}}
            \put(160, 30){\framebox(20, 20){b}}
            \put(180, 30){\framebox(20, 20){a}}
            \uncover<2-2>{
                \put(200, 30){\makebox(100, 20){1 match}}
                \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
                \put(90, 15){\vector(0,1){13}}
                \put(80, 0){\makebox(20, 15){4}}
                \put(120,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
                \put(130, 15){\vector(0,1){13}}
                \put(120, 0){\makebox(20, 15){6}}
            }
            \uncover<3->{
                \put(200, 30){\makebox(100, 20){2 matches}}
            }
            \uncover<3-3>{
                \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
                \put(110, 15){\vector(0,1){13}}
                \put(100, 0){\makebox(20, 15){5}}
                \put(140,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
                \put(150, 15){\vector(0,1){13}}
                \put(140, 0){\makebox(20, 15){7}}
            }
            \uncover<4-4>{
                \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
                \put(130, 15){\vector(0,1){13}}
                \put(120, 0){\makebox(20, 15){6}}
                \put(160,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){b}}}}
                \put(170, 15){\vector(0,1){13}}
                \put(160, 0){\makebox(20, 15){8}}
            }
        \end{picture}
    \end{block}
    \uncover<5->{
        \begin{block}{Result}
            $LCE_s(4,6)=2$
        \end{block}
    }
\end{frame}

\subsection{The \proc{DirectComp} algorithm}
\begin{frame}{Existing Algorithm: \proc{DirectComp}}
    \begin{tabular}{r l}
        Preprocessing & $O(1)$ \\
        Space & $O(1)$ \\
        Query & $O(|\LCE(i,j)|)=O(n)$ \\
        Average query & $O(1)$ \\
        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big)$ \\
    \end{tabular}

@@

\subsection{I/O}
\begin{frame}{The \fprintk\ Algorithm: I/O}
    \begin{itemize}
        \item Original:
        \begin{itemize}
            \item Data structure: $H_\ell[i] = F_{t_\ell}[i]$
            \item Size: $|H_\ell| = n$
            \item I/O: $O\big(k\cdot n^{1/k}\big)$
        \end{itemize}
        \item Cache optimized:
        %Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at 
        \begin{itemize}
            \item Data structure: \\~~$H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1] = F_{t_\ell}[i]$
            \item Size: $|H_\ell| = n+t_\ell$
            \item I/O: $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$
            \begin{itemize}
                \item Best when $k$ is small $\implies$ $n^{1/k}$ is large.
            \end{itemize}
        \end{itemize}
     \end{itemize}
     %$n+t_\ell-1/t_\ell+2$
\end{frame}

@@

\subsection{Practical Results}
\begin{frame}{The \fprintk\ Algorithm: Practical Results}
    Query times of
    \textcolor{red}{\proc{DirectComp}},
    \textcolor{green}{\fprint[2] (cache opt.)},
    \textcolor{blue}{\fprint[3] (not cache opt.)},
    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
    \textcolor{cyan}{\proc{LcpRmq}}
    by string length
    \begin{multicols}{2}{
        \begin{center}
            Average case
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
            \newpage
            Worst case
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-alla.plt}
        \end{center}
    }
    \end{multicols}
\end {frame}

\begin{frame}{The \fprintk\ Algorithm: Practical Results}
    Query times of
    \textcolor{red}{\proc{DirectComp}},
    \textcolor{green}{\fprint[2] (cache opt.)},
    \textcolor{blue}{\fprint[3] (not cache opt.)},
    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
    \textcolor{cyan}{\proc{LcpRmq}}
    by string length
    \begin{multicols}{2}{
        \begin{center}
            Medium LCE size\\
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
            \newpage
            Cache stress\\
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand2.plt}
        \end{center}
    }
    \end{multicols}
\end {frame}

@@

\section{LCE on trees}
\subsection{Compression}
\begin{frame}{LCE on Compressed Strings}
    Goal
    \begin{itemize}
        \item Allow LCE queries without decompressing the string
        \item Using Ziv-Lempel compression (LZ)
    \end{itemize}
    How
    \begin{itemize}
        \item LZ compression represents the string as a tree
        \item An LCE query on a LZ compressed string is a number of LCE queries on a tree
    \end{itemize}
\end{frame}

@@

\begin{frame}{Constant Time String LCE on Heavy Paths}
    \begin{columns}[t]
        \begin{column}{0.7\textwidth}
            \begin{itemize}
                \item How to find the indexes $(i,j)$ in the string:
                \begin{itemize}
                    \item Store an index at each node
                \end{itemize}
                \item How to know then the heavy path splits from the queried path:
                \begin{itemize}
                    \item Store a pointer to the end of the heavy path at each node
                    \item Find NCA of the end of the heavy path and the end of the queried path
                \end{itemize}
                \item How to find a node on the queried path:
                \begin{itemize}
                    \item Use level ancestor
                \end{itemize}
            \end{itemize}
        \end{column}
        \begin{column}{0.3\textwidth}
        \hfill\\
            \includegraphics[width=1\textwidth,page=3]{../doc/tree-fingerprint.pdf}
        \end{column}
    \end{columns}
\end{frame}

\section{Summary}
\begin{frame}{Summary}
    \begin{tabular}{r l l l}
        & \proc{Direct-} & \proc{LcpRmq} / & \\
        & \proc{Comp} & \proc{SuffixNca} & \fprintk \\
        Preprocessing & $O(1)$ & $O(\sortt)$ & $O(k\cdot n+\sortt)$ \\
        Space & $O(1)$ & $O(n)$ & $O(k\cdot n)$\\
        Query & $O(n)$ & $O(1)$ & $O(k\cdot n^{1/k})$ \\
        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
        Query I/O & $O(n/B)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
    \end{tabular}
    \begin{itemize}
        \item In practice, the \fprintk\ algorithm is...
        \begin{itemize}
            \item ...almost as good as \proc{DirectComp} and significantly better than \proc{LcpRmq} in average case
            \item ...significantly better than \proc{DirectComp} but worse than \proc{LcpRmq} in worst case
        \end{itemize}
        \item We can generalize LCE to use on trees:
        \begin{tabular}{r l}
            Preprocessing & $O(\sortt)$ \\
            Space & $O(n)$\\
            Query & $O(\log n)$ \\
        \end{tabular}
    \end{itemize}
\end {frame}

>>>>

\section{Introduction}
\subsection{The LCE Problem}
\begin{frame}{The LCE Problem and the \proc{DirectComp} algorithm}
    \begin{block}{Input}
        \begin{itemize}
            \item $s=abbababba$
            \item $(i,j)=(4,6)$
        \end{itemize}
    \end{block}
    \begin{block}{The \proc{DirectComp} algorithm}
        %\setlength{\unitlength}{1cm}
        \definecolor{matchcolor}{rgb}{.6,1,.6}
        \definecolor{mismatchcolor}{rgb}{1,.6,.6}
        \begin{picture}(300,50)
            \put(0, 30){\makebox(20, 20){$s=$}}
            \put(20, 30){\framebox(20, 20){a}}
            \put(40, 30){\framebox(20, 20){b}}
            \put(60, 30){\framebox(20, 20){b}}
            \put(80, 30){\framebox(20, 20){a}}
            \put(100, 30){\framebox(20, 20){b}}
            \put(120, 30){\framebox(20, 20){a}}
            \put(140, 30){\framebox(20, 20){b}}
            \put(160, 30){\framebox(20, 20){b}}
            \put(180, 30){\framebox(20, 20){a}}
            \uncover<1-2>{
                \put(90, 15){\vector(0,1){13}}
                \put(80, 0){\makebox(20, 15){4}}
                \put(130, 15){\vector(0,1){13}}
                \put(120, 0){\makebox(20, 15){6}}
            }
            \uncover<2-2>{
                \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
                \put(120,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
            }
            \uncover<2-3>{
                \put(200, 30){\makebox(100, 20){1 match}}
            }
            \uncover<3-4>{
                \put(110, 15){\vector(0,1){13}}
                \put(100, 0){\makebox(20, 15){5}}
                \put(150, 15){\vector(0,1){13}}
                \put(140, 0){\makebox(20, 15){7}}
            }
            \uncover<4-4>{
                \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
                \put(140,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
            }
            \uncover<4->{
                \put(200, 30){\makebox(100, 20){2 matches}}
            }
            \uncover<5-6>{
                \put(130, 15){\vector(0,1){13}}
                \put(120, 0){\makebox(20, 15){6}}
                \put(170, 15){\vector(0,1){13}}
                \put(160, 0){\makebox(20, 15){8}}
            }
            \uncover<6-6>{
                \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
                \put(160,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){b}}}}
            }
        \end{picture}
    \end{block}
    \uncover<7->{
        \begin{block}{Result}
            $LCE_s(4,6)=2$
        \end{block}
        \begin{block}{LCE problem}
            Efficiently perform multiple queries $(i,j)$ on a static string $s$
        \end{block}
    }
\end{frame}

\subsection{The \proc{DirectComp} algorithm}
\begin{frame}{Existing Algorithm: \proc{DirectComp}}
    \begin{tabular}{r l}
        Preprocessing & $O(1)$ \\
        Space & $O(1)$ \\
        Query & $O(|\LCE(i,j)|)=O(n)$ \\
        Average query & $O(1)$ \\
        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
    \end{tabular}

@@

\subsection{I/O}
\begin{frame}{The \fprintk\ Algorithm: I/O}
    \begin{itemize}
        \item Original:
        \begin{itemize}
            \item Data structure: $H_\ell[i] = F_{t_\ell}[i]$
            \item Size: $|H_\ell| = n$
            \item I/O: $O\big(k\cdot n^{1/k}\big)$
        \end{itemize}
        \item Cache optimized:
        %Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at 
        \begin{itemize}
            \item Data structure: \\~~$H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1] = F_{t_\ell}[i]$
            \item Size: $|H_\ell| = n+t_\ell$
            \item I/O: $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$
            \begin{itemize}
                \item Best when $k$ is small $\implies$ $n^{1/k}$ is large.
            \end{itemize}
        \end{itemize}
     \end{itemize}
     %$n+t_\ell-1/t_\ell+2$
    \begin{center}
        \includegraphics[width=0.7\textwidth,page=2]{../doc/fingerprint.pdf}\\
    \end{center}
\end{frame}

@@

\subsection{Practical Results}
\begin{frame}{Practical Results}
    Query times of
    \textcolor{red}{\proc{DirectComp}},
    \textcolor{green}{\fprint[2] (cache opt.)},
    \textcolor{blue}{\fprint[3] (not cache opt.)},
    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
    \textcolor{cyan}{\proc{LcpRmq}}
    by string length
    \begin{multicols}{2}{
        \begin{center}
            Average case
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
            \newpage
            Worst case
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-alla.plt}
        \end{center}
    }
    \end{multicols}
\end {frame}

\begin{frame}{Practical Results}
    Query times of
    \textcolor{red}{\proc{DirectComp}},
    \textcolor{green}{\fprint[2] (cache opt.)},
    \textcolor{blue}{\fprint[3] (not cache opt.)},
    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
    \textcolor{cyan}{\proc{LcpRmq}}
    by string length
    \begin{center}
        Medium LCE size\\
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
    \end{center}
\end {frame}

\begin{frame}{Cache Optimization, Practical Results}
    Is I/O optimization good in practice?
    \begin{itemize}
        \item Pro: better cache efficiency
            \begin{itemize}
                \item Best for small $k$, no change for $k=\logceil$
            \end{itemize}
        \item Con: Calculating memory addresses is more complicated
            \begin{itemize}
                \item $((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1$ vs.\ $i$
            \end{itemize}
    \end{itemize}
\end{frame}

@@

\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
    Query times of
    \textcolor{red}{\proc{DirectComp}},
    \textcolor{green}{\fprint[2] (cache opt.)},
    \textcolor{blue}{\fprint[3] (not cache opt.)},
    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
    \textcolor{cyan}{\proc{LcpRmq}}
    by string length
    \begin{multicols}{2}{
        \begin{center}
            Average case
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
            \newpage
            Cache stress\\
            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand2.plt}
        \end{center}
    }
    \end{multicols}
\end {frame}

\begin{comment}
\section{LCE on trees}
\subsection{Compression}
\begin{frame}{LCE on Compressed Strings}
    Goal
    \begin{itemize}
        \item Allow LCE queries without decompressing the string
        \item Using Ziv-Lempel compression (LZ)
    \end{itemize}
    How
    \begin{itemize}
        \item LZ compression represents the string as a tree
        \item An LCE query on a LZ compressed string is a number of LCE queries on a tree
    \end{itemize}
\end{frame}

@@

\begin{frame}{Constant Time String LCE on Heavy Paths}
    \begin{columns}[t]
        \begin{column}{0.7\textwidth}
            \begin{itemize}
                \item How to find the indexes $(i,j)$ in the string:
                \begin{itemize}
                    \item Store an index at each node
                \end{itemize}
                \item How to know then the heavy path splits from the queried path:
                \begin{itemize}
                    \item Store a pointer to the end of the heavy path at each node
                    \item Find NCA of the end of the heavy path and the end of the queried path
                \end{itemize}
                \item How to find a node on the queried path:
                \begin{itemize}
                    \item Use level ancestor
                \end{itemize}
            \end{itemize}
        \end{column}
        \begin{column}{0.3\textwidth}
        \hfill\\
            \includegraphics[width=1\textwidth,page=3]{../doc/tree-fingerprint.pdf}
        \end{column}
    \end{columns}
\end{frame}
\end{comment}

\section{Summary}
\begin{frame}{Summary}
    \begin{tabular}{r l l l}
        & \proc{Direct-} & \proc{LcpRmq} / & \\
        & \proc{Comp} & \proc{SuffixNca} & \fprintk \\
        Preprocessing & $O(1)$ & $O(\sortt)$ & $O(k\cdot n+\sortt)$ \\
        Space & $O(1)$ & $O(n)$ & $O(k\cdot n)$\\
        Query & $O(n)$ & $O(1)$ & $O(k\cdot n^{1/k})$ \\
        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
        Query I/O & $O\big(\frac{n}{B}\big)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
    \end{tabular}
    \vspace{1em}
    \begin{itemize}
        \item In practice, the \fprintk\ algorithm is...
        \begin{itemize}
            \item ...almost as good as \proc{DirectComp} and significantly better than \proc{LcpRmq} in average case
            \item ...significantly better than \proc{DirectComp} but worse than \proc{LcpRmq} in worst case
        \end{itemize}
        \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
        %\item We can generalize LCE to use on trees:
        %\begin{tabular}{r l}
        %    Preprocessing & $O(\sortt)$ \\
        %    Space & $O(n)$\\
        %    Query & $O(\log n)$ \\
        %\end{tabular}
    \end{itemize}
\end {frame}
